---
title: "A2Q3"
author: "Lily Li"
date: "2025-06-11"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Q3

```{r}
library(cancensus)

#all_vecs = list_census_vectors("CA21")
# Total English         v_CA21_2209
# Total French          v_CA21_2212
# Full time Full year   v_CA21_6525
# Average wks worked    v_CA21_6531
# Median after-tax inc  v_CA21_986
# Median Total inc      v_CA21_983


all_regions = list_census_regions("CA21") # 2021 census
csd_all = list(CSD = all_regions$region)
all_regions = data.frame(all_regions)

thislevel = "CSD"
census_data = get_census(dataset = "CA21",
                         regions = csd_all,
                         vectors = c("v_CA21_2209", 
                                     "v_CA21_2212", 
                                     "v_CA21_983", 
                                     "v_CA21_986", 
                                     "v_CA21_6525", 
                                     "v_CA21_6531"),
                         level = thislevel)

names(census_data)[ncol(census_data) - 5 ] = "English_Speakers"
names(census_data)[ncol(census_data) - 4 ] = "French_Speakers"
names(census_data)[ncol(census_data) - 3 ] = "Median_Before_Tax"
names(census_data)[ncol(census_data) - 2 ] = "Median_After_Tax"
names(census_data)[ncol(census_data) - 1 ] = "FTime_FYear_Workers"
names(census_data)[ncol(census_data) - 0 ] = "Avg_Weeks_Worked"


census_data = data.frame(census_data)
census_data = subset(census_data, Population > 0 & !is.na(Avg_Weeks_Worked) 
                     & !is.na(Median_After_Tax))
filename = paste0("Stat341_S25_A2_Census_Data_",thislevel,".csv")
write.csv(census_data, filename, row.names=FALSE)

```




# (a)


```{r}
library(quantreg)

# for testing, i'll use Median_Before_Tax to predict Median_After_Tax
my_data <- data.frame(
  x = census_data$Median_Before_Tax,
  y = census_data$Median_After_Tax
)

```


```{r}

rho_lm_fn <- function(x, data) {
  alpha <- x[1]
  beta <- x[2]
  y_hat <- alpha + beta * data[[1]]
  residuals <- data[[2]] - y_hat
  sum(residuals^2)
}

```

# (b)

```{r}

opt_lm <- optim(c(0, 0), rho_lm_fn, data = my_data)
print(opt_lm$par)

lm_model <- lm(y ~ x, data = my_data)
print(coef(lm_model))
```

# (c)
```{r}
rho_qr_fn <- function(x, tau, data) {
  alpha <- x[1]
  beta <- x[2]
  residuals <- data[[2]] - (alpha + beta * data[[1]])
  sum(ifelse(residuals >= 0, tau * residuals, (tau - 1) * residuals))
}

```

# (d)

```{r}
taus <- c(0.1, 0.25, 0.5, 0.75, 0.9)

results <- data.frame(
  Tau = taus,
  Alpha_optim = NA, Beta_optim = NA,
  Alpha_rq = NA, Beta_rq = NA
)

for (i in seq_along(taus)) {
  tau <- taus[i]
  
  opt_result <- optim(c(0, 0), rho_qr_fn, tau = tau, data = my_data)
  rq_model <- rq(y ~ x, tau = tau, data = my_data)
  
  results$Alpha_optim[i] <- opt_result$par[1]
  results$Beta_optim[i] <- opt_result$par[2]
  results$Alpha_rq[i] <- coef(rq_model)[1]
  results$Beta_rq[i] <- coef(rq_model)[2]
  
}

print(results)

```


# (e)

The slope estimates from both optim() and rq() are generally close, indicating the generalized optimizer approximates the quantile regression well. However, there is a significant discrepancy in the intercept estimates, likely due to scaling or sign-handling in the asymmetric loss function. This suggests that while optim() is flexible, it may require careful setup, and domain-specific functions like rq() are more stable and robust for this type of modeling.

