---
title: "A4Q1"
author: "Lily Li"
date: "2025-07-28"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Q1
# (a)


```{r}
df = read.csv("Stat341_S25_LinkedIn_Dataset.csv")
head(df)
dim(df)
```

```{r}

x <- df$Reacts
y <- df$Impressions

r_result <- cor.test(x,y, method = "pearson")
r <- r_result$estimate

r

n <- length(x)

t_obs <- r * sqrt(n-2) / sqrt(1-r^2)
p_val <- 2 * pt(-abs(t_obs), df = n-2)

p_val
```


The resulting p-value is less than 0.05, providing sufficient evidence to reject the null hypothesis. This suggests that there is a statistically significant linear correlation between Impressions and Reacts in the dataset.


# (b)

```{r}

transforms <- list(
  identity = function(v) v,
  log = function(v) log(v + 1),
  sqrt = function(v) sqrt(v),
  inv = function(v) -1/(v + 1)
)

results <- data.frame(
  transformation = character(), r = numeric(), p = numeric()
)

for (name in names(transforms)) {
  tx <- transforms[[name]](x)
  ty <- transforms[[name]](y)
  r_test <- cor.test(tx,ty, method = "pearson")
  r_val <- r_test$estimate
  t_stat <- r_val * sqrt(n-2) / sqrt(1-r_val^2)
  p_val <- 2 * pt(-abs(t_stat), df = n-2)
  results <- rbind(results, data.frame(transformation = name, r=r_val, t=t_stat, p=p_val))

}

print(results)

```



# (c)

```{r}
randomPairing = function(x,y)
{
  xrand = sample(x)
  yrand = sample(y)
  
  return(list(x=xrand, y=yrand))
}
```


```{r}

rp <- randomPairing(x,y)
par(mfrow = c(2,2))

hist(x, main = "Original Reacts (x)")
hist(y, main = "Original Impressions (y)")
hist(rp$x, main = "Random Pairing x")
hist(rp$y, main = "Random Pairing y")


```


For Original React, the distribution right-skewed, with a cluster around smaller values and a few higher outliers, like around 15. In such case, log(x+1) or a square root transformation would be helpful in reducing skewness and stabilize variance. 

For Original Impression, it is strongly right skew, with most of the values centered around 1000, and a few larger outliers, like 5000. This suggests a suitability of log(y+1) here to compress the large values and make the distribution more symmetric.

For both random pairing x and y, their distributions look nearly identical to the originals. 

# (d)

```{r}

M <- 5000

t_scores <- replicate(M, {
  
  rp <- randomPairing(x, y)
  rp_test <- cor.test(rp$x, rp$y, method = "pearson")
  rp_val <- rp_test$estimate
  rp_val * sqrt(n - 2) / sqrt(1 - rp_val^2)
  
})

hist(t_scores, breaks = 50, main = "Random Pairing t-scores")
abline(v = t_obs, col = "red", lwd = 3)


```


# (e)

```{r}
p_empirical <- mean(abs(t_scores) >= abs(t_obs))

p_empirical

```

# (f)

```{r}
get_pval <- function(x,y) {
  r <- cor(x,y)
  t <- r * sqrt(n - 2) / sqrt(1 - r^2)
  2 * pt(-abs(t), df = n - 2)
}

pvals_matrix <- matrix(NA, nrow = M, ncol = 4)

for (i in 1:M) {
  
  rp <- randomPairing(x, y)
  for (j in 1:4) {
    tx <- transforms[[j]](rp$x)
    ty <- transforms[[j]](rp$y)
    pvals_matrix[i, j] <- get_pval(tx, ty)
  }
}

plot(pvals_matrix[1:500, 1], pvals_matrix[1:500, 2],
     xlab = "Untransformed p-values", ylab = "Log-transformed p-values",
     main = "p-values: Untransformed vs Log", pch = 20)
```


# (g)

```{r}

observed_pvals <- sapply(transforms, function(f) {
  tx <- f(x)
  ty <- f(y)
  get_pval(tx, ty)
})

observed_min <- min(observed_pvals)

min_pvals_random <- apply(pvals_matrix, 1, min)
p_adj <- mean(min_pvals_random <= observed_min)

p_adj
```

Among the 5000 random pairings, none of the transformations for each pairing produced a smaller minimum p-value, thereby the proportion is zero. This indicates a strong evidence against the null hypothesis of independence, even across multiple transformations. 