---
title: "A1Q2"
author: "Lily Li"
date: "2025-05-15"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Q2


```{r}
library(cancensus)
library(readr)
library(ggplot2)
library(dplyr)
all_regions = list_census_regions("CA21") # 2021 census
rural_on_regions = subset(all_regions,
                          level=="CSD" &  # census subdivisions
                            PR_UID == 35 & # in Ontario
                            is.na(CMA_UID)) # not inside a metropolitan area

csd_list = list(CSD = rural_on_regions$region)

census_data = get_census(dataset = "CA21",
                         regions = csd_list,
                         level = "CSD")

names(census_data)[4] = "Area"
head(census_data)
```

# (b)

```{r}
cor(census_data$Area, census_data$Population, use = "complete.obs")
```


# (c)

```{r}
ggplot(census_data, aes(x = Area, y = Population)) + geom_point(alpha = 0.6) +
  labs(title = "Population vs Area of Rural CSDs", 
       x = "Land Area km²", y = "Population") + 
  theme_minimal()
```

From the scatterplot above, it is clear that most data points are densely clustered near the origin, while a few extreme outliers with very large land areas stretch the x-axis. This uneven spread makes it difficult to observe any meaningful relationship in the raw scale. Applying a **logarithmic transformation** to both area and population would help by **compressing the large values** and **spreading out the smaller, clustered points**, thereby making the overall relationship more linear and likely increasing the correlation coefficient. 


# (d)

```{r}
transformations = list(log = function (x) log(x + 1), sqrt = function(x) sqrt(x),
                       inv = function(x) -1 / (x + 1), square = function(x) x^2,
                       cube = function(x) x^3)

results = data.frame(Area_transformed = character(0),
                     Population_transformed = character(0),
                     Correlation = numeric(0))
for(area_name in names(transformations)) {
  for(pop_name in names(transformations)) {
    try({
      tx = transformations[[area_name]](census_data$Area)
      ty = transformations[[pop_name]](census_data$Population)
      cor_val = cor(tx, ty, use = "complete.obs")
      results = rbind(results, data.frame(Area_transformed = area_name, 
                                          Population_transformed = pop_name, 
                                          Correlation = cor_val))
    }, silent = TRUE)
  }
}

results <- results %>% mutate(abs_corr = abs(Correlation)) %>% 
  arrange(desc(abs_corr))

print(results)
```


Among all tested combinations of power transformations, the **logarithmic transformation applied to both area and population** resulted in the strongest linear correlation. The corresponding Pearson correlation coefficient was **0.5098**, indicating a moderate positive linear relationship after the log-log transformation.

