---
title: "A3Q2"
author: "Lily Li"
date: "2025-07-05"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Q2
# (a)

```{r}
mlb_bs <- read.csv("MLB_Batting_Boxscores_2024-04.csv")
mlb_visitor <- mlb_bs[1:9, ]
mlb_home <- mlb_bs[10:18, ]
```

```{r}
yV = sort(mlb_visitor$total_bases) 
yH = sort(mlb_home$total_bases)

x = ((1:length(yV)) - 0.5)/length(yV)

par(mfrow = c(1,3))


plot(x, yV, xlab="Quantile", ylab="Bases", pch=16,
las=1, ylim=c(0,max(yV, yH)), main="Visitor")
plot(x, yH, xlab="Quantile", ylab="Bases", pch=16,
las=1, col = "Red", ylim=c(0,max(yV, yH)), main="Home")

plot(x, yV, xlab="Quantile", ylab="Bases", pch=16,
las=1, ylim=c(0,max(yV, yH)), main="Both")
points(x, yH, col="Red", pch=16)

par(mfrow = c(1,1))

```


# (b)

H0: The total_base distribution for visitors and home batters arefrom the same population. 

# (c)

```{r}
mean_V <- mean(mlb_visitor$total_bases)
mean_H <- mean(mlb_home$total_bases)

D_obs <- abs(mean_V - mean_H)

D_obs
```

# (d)
```{r}

mixRandomly <- function(pop) {  
  
  pop1 <- pop$pop1
  n_pop1 <- nrow(pop1)
  pop2 <- pop$pop2
  n_pop2 <- nrow(pop2)
  mix <- rbind(pop1,pop2)
  select4pop1 <- sample(1:(n_pop1 + n_pop2),
  n_pop1,
  replace = FALSE)
  new_pop1 <- mix[select4pop1,]
  new_pop2 <- mix[-select4pop1,]
  list(pop1=new_pop1, pop2=new_pop2)

}
```

```{r}
pop_list <- list(pop1 = mlb_visitor, pop2 = mlb_home)

M <- 5000
D_star <- numeric(M)

set.seed(341)
for(i in 1:M) {
  mixed <- mixRandomly(pop_list)
  yV_star <- mixed$pop1$total_bases
  yH_star <- mixed$pop2$total_bases
  D_star[i] <- abs(mean(yV_star) - mean(yH_star))
}

hist(D_star, breaks=30, main="Permutation Distribution of D", xlab="Discrepancy D")
abline(v=D_obs, col="red", lwd=2)

```

# (e)

```{r}
p_value_approx <- mean(D_star >= D_obs)
p_value_approx
```

# (f)

Since the p-value exceeds 0.05, we do not have sufficient evidence to reject H0. We conclude that there is no significant difference in total bases between the visitor and home player populations.


# (g)
```{r}
combos <- combn(18, 9)
PV_10000 <- combos[,10000]
PH_10000 <- setdiff(1:18, PV_10000)

PV_10000
PH_10000

```

# (h)

```{r}
D_all <- numeric(ncol(combos))

y_all <- mlb_bs$total_bases[1:18]

for (i in 1:ncol(combos)) {
  PV_idx <- combos[,i]
  PH_idx <- setdiff(1:18, PV_idx)
  D_all[i] <- abs(mean(y_all[PV_idx]) - mean(y_all[PH_idx]))
}
  p_value_exact <- mean(D_all >= D_obs)
  p_value_exact
```

# (i)

There is not substantial difference between the results from part (h) and here, this is not surprising because 5000 permutations is large enough to approximate the exact distribution closely. 
